Rising from the Ashes II: The Bar-driven Abundance Bimodality of the Milky Way

### 2025-02-18 paper/ms.tex
Update alpha and obs errors

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{392276_alpha.pdf}
  \caption{\textbf{Bimodality in the abundance plane is linked to distinct epochs separated by quiescence in simulation.} The upper panels show \MgFe{} as a function of age for our subhalo in TNG. The colors indicate stellar populations at fixed values of \FeH{}, which are the same as in Figure~\ref{fig:fig1}. A gap in the relation occurs at an age of approximately $10.6\,\Gyr$, which we indicate with a vertical dashed line. The effect of the $\alpha$-enhancement is clear, as it separates the stars that form before and after this gap in ages (star particles which formed before $z=1.5$ are $\alpha$-enhanced, which occurs at an age of $\sim9.5\,\Gyr$). The lower panels show on the left the Milky Way and on the center and right the data from TNG but with $12.5\%$ age errors and $0.01\,\dex$ errors in \MgFe{}. When the simulations are given these errors, we see that the before and after star particles smear such that the two populations significantly overlap in ages. In the $\alpha$-enhanced galaxy, two populations emerge in each bin which overlapped in the fiducial distribution. This feature more closely resembles the Milky Way, which displays such populations where the bimodality is strongest -- $\FeH=-0.5$ (blue) and $-0.25$ (orange).}
  \label{fig:alpha}
\end{figure*}

The abundance distributions shown in Figure~\ref{fig:fig1} can be better understood by examining the evolution of \MgFe{} with time of the individual stars/star particles. In the upper panels of Figure~\ref{fig:alpha} we show the true distribution of \MgFe{} as a function of time for the fiducial galaxy in the middle and for the post-processed, $\alpha$-enhanced subhalo to the right. We use age instead of formation time in order to better facilitate comparisons to observations. These panels show 2D histograms, with a logarithmic colormap normalized to the maximum of the plot. To prevent overlap, the values of \MgFe{} are given offsets of $0$, $-0.5$, $-1$, and $-1.5$, in order of increasing \FeH{}.

In our simulated galaxy, there is an age gap at $\sim10.6\,\Gyr$, which we mark with a vertical dashed line in the upper row. Star particles older than this line have a much clearer gradient in time with \MgFe{} than stars that form after. In the $\FeH=-0.25$ bin, star particles which form directly after this line have a slightly reduced \MgFe{} than stars which form a short time later.

In the lower row, the left panel displays data from the Milky Way, while the middle and right panels show the fiducial and $\alpha$-enhanced galaxy simulations, respectively. Gaussian errors of $12.5\%$ in age and $0.01\,dex$ in \MgFe{} were introduced in the simulations which align with the observational uncertainties in the APOKASC-3 and APOGEE datasets (see Appendix~\ref{app:obs_err} for detailed plots of observational errors).

While the error in \MgFe{} is insignificant, the age error ($1.25\,\Gyr$ at $10\,\Gyr$) blurs the distribution, particularly around the dashed line that marks the transition between sequences. Despite this, the $\alpha$-enhanced galaxy still shows two distinct populations, although their ages now overlap more significantly.

The Milky Way distribution (lower left panel) exhibits some similarities to the $\alpha$-enhanced galaxy. Specifically, in the $\FeH=-0.5$ and $-0.25$ bins (blue and orange, respectively), there appear to be two populations that overlap in age but remain distinct in \MgFe{}. These are the bins where the bimodality is most pronounced (as seen in the upper left panel of Figure~\ref{fig:fig1}). In the Milky Way there also appears to be a population of $\alpha$-rich stars down to low ages (see discussion in Section~\ref{ssec:compare_obs}).

@@

\subsection{Direct Comparison to Observations}\label{ssec:compare_obs}
The lower left panel of Figure~\ref{fig:alpha} shows the \MgFe{} vs. age of stars in bins of \FeH{} which pass our solar neighborhood selection and are present in the APOKASC-3 catalog. In the bins where the bimodality is strongest (blue and orange, $\FeH=-0.5$ and $-0.25$, respectively), we see that there is a sort of two tiered distribution with significant overlap in age. In the galaxy with age and abundance errors shown in the lower right panel, a similar distribution can be seen in all \FeH{} bins. An examination of the true distribution in the galaxy (upper right panel), we see that the two distributions are in fact very cleanly separated in age.

@@

We argue that the reduction in the SFR has two effects which lead to the $\alpha$-bimodality. First, a lower production of Mg relative to Fe is the natural result of a lack of Type~II relative to Type~Ia SNe. The latter remain high due to their longer timescales, expediting a drop in \MgFe{}. Second, the lack of SF means that stars at \MgFe{} intermediate between the high- and low-$\alpha$ sequences never form. This argument is the same as presented in Paper~I and is consistent with the one-zone chemical evolution models of \citet{2020MNRAS.498.1364J}. When observational errors are added to the subhalo's abundance vs. age distribution, we show that it is broadly consistent with the distribution observed in the Milky Way (Figure~\ref{fig:alpha} and Section~\ref{ssec:compare_obs}).

@@

\section{Observational Errors}\label{app:obs_err}
In Figure~\ref{fig:alpha}, we assumed observational errors of $12.5\%$ in age and $0.01\,\dex$ in \MgFe{}. In Figure~\ref{fig:obs_err}, we plot the quoted observational errors of the APOKASC-3 (left) and ASPCAP (right) datasets, showing both \FeH{} and \MgFe{} (blue and orange, respectively). We show our $12.5\%$ age error and $0.01\,\dex$ abundance error assumptions as black lines. For the age error, we used the maximum of the upper and lower estimates from \citet{2018ApJS..239...32P}. As a dashed line we show our $25\%$ age error cut for stars plotted in Figure~\ref{fig:obs_err}. Our assumed errors are generally consistent with the observed. Our error in \MgFe{} is a bit smaller than in ASPCAP, but the age estimates are by far the more constraining of the two.

\begin{figure*}
  \centering
  \includegraphics[width=\columnwidth]{obs_error.pdf}
  \caption{The observational errors of the APOKASC-3 (left) and ASPCAP dataset (right). We show, on the left, a line indicating a $12.5\%$ error in observed age and on the right a vertical line indicating a $0.01\,\dex$ error. On the left, a dashed line indicates the $25\%$ error cut used for inclusion in Figure~\ref{fig:alpha}.}
  \label{fig:obs_err}
\end{figure*}

>>>>

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{392276_alpha.pdf}
  \caption{\textbf{Bimodality in the abundance plane is linked to distinct epochs separated by quiescence in simulation.} The upper row shows \MgFe{} as a function of age for our subhalo in TNG. The colors indicate stellar populations at fixed values of \FeH{}, which are the same as in Figure~\ref{fig:fig1}. A gap in the relation occurs at an age of approximately $10.6\,\Gyr$, which we indicate with a vertical dashed line. The effect of the $\alpha$-enhancement is clear, as it separates the stars that form before and after this gap in ages (star particles which formed before $z=1.5$ are $\alpha$-enhanced, which occurs at an age of $\sim9.5\,\Gyr$). \edit1{The middle row shows the same TNG and $\alpha$-enhanced TNG data, but with added uncertainties of $12.5\%$ in age and $0.015\,\dex$ in \MgFe{}.} When given these errors, the before and after star particles smear such that the two populations significantly overlap in ages. \edit1{The lower row shows on the left the Milky Way data and on the right the same TNG data with artificial errors but subsampled to the same number of stars older than $5\,\Gyr$ as in the observations (because the simulation sample has almost no star particles younger than $5\,\Gyr$). The limited sample size of the observations makes a direct comparison difficult.}}
  \label{fig:alpha}
\end{figure*}

The abundance distributions shown in Figure~\ref{fig:fig1} can be better understood by examining the evolution of \MgFe{} with time of the individual stars/star particles. In the upper panels of Figure~\ref{fig:alpha} we show the true distribution of \MgFe{} as a function of time for the fiducial galaxy in the middle and for the post-processed, $\alpha$-enhanced subhalo to the right. We use age instead of formation time in order to better facilitate comparisons to observations. These panels show 2D histograms, with a logarithmic colormap normalized to the maximum of the plot. To prevent overlap, the values of \MgFe{} are given offsets of $0$, $-0.5$, $-1$, and $-1.5\,\dex$, in order of increasing \FeH{}.

In our simulated galaxy, there is an age gap at $\sim10.6\,\Gyr$, which we mark with a vertical dashed line in the upper row. Star particles older than this line have a much clearer gradient in time with \MgFe{} than stars that form after, \edit1{even in the fiducial case}. In the $\FeH=-0.25$ bin, star particles which form directly after this line have a slightly reduced \MgFe{} than stars which form a short time later.

\edit1{In the middle row, the center and right panels show the simulated galaxies with Gaussian errors of $12.5\%$ in age and $0.015\,\dex$ in \MgFe{}, aligning with the observational uncertainties in the APOKASC-3 and APOGEE datasets (see Appendix~\ref{app:obs_err}). The error in \MgFe{} is insignificant, but the age error ($1.25\,\Gyr$ at $10\,\Gyr$) significantly blurs the distribution, particularly across the dashed line that marks the transition between sequences. Despite this, the $\alpha$-enhanced galaxy still shows two distinct populations, although their ages now overlap more significantly.}

\edit1{The lower row compares Milky Way data (left) with simulations of fiducial (middle) and $\alpha$-enhanced (right) galaxies. Here, the simulations have been subsampled to match the observed sample size of stars older than $5\,\Gyr$ in each metallicity bin. We match the sample size only to old stars because our simulated sample has almost no star particles younger than $5\,\Gyr$. When subsampled with observational errors, the $\FeH=-0.5$ bin in the simulation (orange right panel) very faintly shows a hint of two stacked distributions which might also be present in the $\FeH=-0.25$ bin in the data (orange left panel). The limited sample size makes it impossible to draw any strong conclusions.}

@@

\subsection{Direct Comparison to Observations}\label{ssec:compare_obs}
\edit1{We attempted a direct comparison between our TNG galaxy and the Milky Way in the lower row of Figure~\ref{fig:alpha} (see Section~\ref{ssec:alpha_time}). In the simulations, the presence of the bimodality is obvious from a stacked distribution. However, this distribution is not obvious in the observations, primarily because the sample size of observations at old ages where the bimodality is present ($>5\,\Gyr$) is quite small. Nonetheless, the presence of a clean gap in stellar ages between the high- and low-$\alpha$ is completely washed out by the age uncertainties.}

@@

We argue that the reduction in the SFR has two effects which lead to the $\alpha$-bimodality. First, a lower production of Mg relative to Fe is the natural result of a lack of Type~II relative to Type~Ia SNe. The latter remain high due to their longer timescales, expediting a drop in \MgFe{}. Second, the lack of SF means that stars at \MgFe{} intermediate between the high- and low-$\alpha$ sequences never form. This argument is the same as presented in Paper~I and is consistent with the one-zone chemical evolution models of \citet{2020MNRAS.498.1364J}. \edit1{Because of observational errors in age, it is not possible to make a direct comparison between the simulated galaxy and datae (Figure~\ref{fig:alpha} and Section~\ref{ssec:compare_obs}).}

@@

\section{Observational Errors}\label{app:obs_err}
\edit1{In Figure~\ref{fig:alpha}, we assumed observational errors of $12.5\%$ in age and $0.015\,\dex$ in \MgFe{}. In Figure~\ref{fig:obs_err}, we plot the quoted observational errors of the APOKASC-3 (left) and ASPCAP (right) datasets, showing both \FeH{} and \MgFe{} (blue and orange, respectively). We show our $12.5\%$ age error as a black line in the left panel. We take the age error to be the maximum of the upper and lower errors from \citet{2018ApJS..239...32P}. In the right panel, we show blue and orange vertical lines at $0.01$ and $0.015\,\dex$ for \FeH{} and \MgFe{}, respectively. These are approximately the 99th percentiles of each error distribution. As a dashed line we show our $25\%$ age error cut for stars plotted in Figure~\ref{fig:obs_err}. Our assumed errors are generally consistent with the errors quoted in the dataset, with a conservative estimate for the abundance errors.}

\begin{figure*}
  \centering
  \includegraphics[width=\columnwidth]{obs_error.pdf}
  \caption{The observational errors of \edit1{astroseismic ages from APOKASC-3} (left) and \edit1{abundances from ASPCAP DR17} (right). We show, on the left, a line indicating a $12.5\%$ error in observed age and on the right a vertical line indicating a \edit1{$0.01$ and $0.015\,\dex$ error in \FeH{} and \MgFe{}, respectively}. On the left, a dashed line indicates the $25\%$ error cut used for inclusion in Figure~\ref{fig:alpha}.}
  \label{fig:obs_err}
\end{figure*}
